% Manual references for CIKM.tex
\begin{thebibliography}{20}

\bibitem{tamm2024}
Y.-M. Tamm and A. Aljanaki, ``Comparative analysis of pretrained audio representations in music recommender systems,'' in \emph{Proceedings of the 18th ACM Conference on Recommender Systems (RecSys '24)}, Bari, Italy, Oct. 2024, pp. 1088--1092. doi: 10.1145/3640457.3688172.

\bibitem{zaman2023}
K.~Zaman, M.~Sah, C.~Direkoglu, and M.~Unoki, ``A survey of audio classification using deep learning,'' \emph{IEEE Access}, vol.~11, pp. 102\,345--102\,381, 2023.

\bibitem{jiang2024}
X.~Jiang, Y.~Zhang, G.~Lin, and L.~Yu, ``Music emotion recognition based on deep learning: A review,'' \emph{IEEE Access}, vol.~12, pp. 157\,716--157\,745, 2024.

\bibitem{lin2024}
P.-C.~Lin \emph{et~al.}, ``Prototype for a personalized music recommendation system based on TL-CNN-GRU model with 10-fold cross-validation,'' 2024.

\bibitem{pourmoazemi2024}
N.~Pourmoazemi and S.~Maleki, ``A music recommender system based on compact convolutional transformers,'' \emph{Expert Systems with Applications}, vol.~255, p. 124473, 2024.

\bibitem{gong2021ast}
Y.~Gong, Y.-A. Chung, and J.~Glass, ``Ast: Audio spectrogram transformer,'' \emph{arXiv preprint arXiv:2104.01778}, 2021.

\bibitem{krichene2020}
W.~Krichene and S.~Rendle, ``On sampled metrics for item recommendation,'' in \emph{Proceedings of the 26th ACM SIGKDD Conference on Knowledge Discovery and Data Mining (KDD '20)}.\hskip 1em plus 0.5em minus 0.4em\relax Virtual Event, CA, USA: ACM, 2020, pp. 1748--1757.

\bibitem{canamares2020}
R.~Ca\~namares and P.~Castells, ``On target item sampling in offline recommender system evaluation,'' in \emph{Proceedings of the 14th ACM Conference on Recommender Systems (RecSys '20)}.\hskip 1em plus 0.5em minus 0.4em\relax Virtual Event, Brazil: ACM, 2020, pp. 259--268.

\bibitem{porcaro2022}
L.~Porcaro, E.~G\'omez, and C.~Castillo,
``Assessing the impact of music recommendation diversity on listeners: A longitudinal study,''
\emph{ACM Transactions on Intelligent Systems and Technology}, 2020.

\bibitem{lin2025}
J.~Lin, S.~Huang, and Y.~Zhang,
``Deep neural network-based music user preference modeling, accurate recommendation, and IoT-enabled personalization,''
2025.

\bibitem{abbattista2024}
D.~Abbattista, V.~W. Anelli, T.~Di~Noia, C.~Macdonald, and A.~Petrov, ``Enhancing sequential music recommendation with personalized popularity awareness,'' in \emph{Proceedings of the 18th ACM Conference on Recommender Systems (RecSys '24)}, Bari, Italy, Oct. 2024.

\bibitem{schedl2021}
M.~Schedl, C.~Bauer, and W.~Reisinger, ``Listener modeling and context-aware music recommendation based on country archetypes,'' \emph{Frontiers in Artificial Intelligence}, vol.~3, 2021.

\bibitem{eerola2011}
T.~Eerola and J.~K. Vuoskoski, ``A comparison of the discrete and dimensional models of emotion in music,'' \emph{Psychology of Music}, vol.~39, no.~1, pp. 18--49, 2011.

\bibitem{urbano2013}
J.~Urbano, M.~Schedl, and X.~Serra, ``Evaluation in music information retrieval,'' \emph{Journal of Intelligent Information Systems}, 2013.

\bibitem{kong2020panns}
Q.~Kong, Y.~Cao, T.~I. Li, Y.~Wang, and M.~D. Plumbley,
``PANNs: Large-scale pretrained audio neural networks for audio pattern recognition,''
\emph{IEEE/ACM Transactions on Audio, Speech, and Language Processing},
vol.~28, pp. 2880--2894, 2020.

\bibitem{li2023mert}
Y.~Li, X.~Chen, and others, ``MERT: Acoustic Music Understanding Model with Large-Scale Self-supervised Training,'' \emph{arXiv preprint arXiv:2306.00107}, 2023.

\bibitem{schedl2018}
M.~Schedl, H.~Zamani, C.-W.~Chen, Y.~Deldjoo, and M.~Elahi,
``Current challenges and visions in music recommender systems research,''
\emph{International Journal of Multimedia Information Retrieval},
vol.~7, pp.~95--116, 2018.
doi: 10.1007/s13735-018-0154-2.

\bibitem{deldjoo2024}
Y.~Deldjoo, M.~Schedl, and P.~Knees,
``Content-driven music recommendation: Evolution, state of the art, and challenges,''
\emph{Computer Science Review},
vol.~51, p.~100618, 2024.
doi: 10.1016/j.cosrev.2024.100618.

\bibitem{sun2019bert4rec}
F.~Sun, J.~Liu, J.~Wu, C.~Pei, X.~Lin, W.~Ou, and P.~Jiang,
``BERT4Rec: Sequential recommendation with bidirectional encoder representations from transformer,''
in \emph{Proceedings of the 28th ACM International Conference on Information and Knowledge Management (CIKM '19)},
Beijing, China, 2019, pp. 1441--1450.

\bibitem{yilmaz2008}
E.~Yilmaz and J.~A. Aslam,
``Estimating average precision with incomplete and imperfect judgments,''
in \emph{Proceedings of the 17th ACM Conference on Information and Knowledge Management (CIKM '08)},
Napa Valley, CA, USA, 2008, pp. 1021--1030.

\end{thebibliography}
